Large-scale Omics Experiments are not Conducive to Causal Inference

### 2015-09-07 write_up.Rmd
olgas intro

Causal inference – the task of uncovering regulatory relationships between components of biomolecular pathways and networks - is a primary goal of omics datasets in molecular biology studies. Generally, observing statistical association between network components does not demonstrate causality. However, we can safely use association observed in measurements generated by a sound experimental design as evidence for the presence of a causal relation.

High-throughput omics technologies enable large-scale experiments, with simultaneous measurements of multiple cellular components in a sample. The huge amount of data generated paradoxically make the task of causal inference much more difficult because, when many examining pairwise associations between measured features, some will appear significant just by chance. In other words, large-scale experiments are more prone to hiding the true signal and generating spurious associations, leading to increased false positive conclusions that causal relationships are present. This problem can be addressed by improving experimental design and refining the biological question. Below, we describe the process of elucidating statistical associations from data and suggest practical approaches for analysis of large scale datasets.

>>>>

Causal inference – the task of uncovering regulatory relationships between components of biomolecular pathways and networks - is a primary goal of omics datasets in molecular biology studies. Large-scale experiments are fishing expeditions, where we look for associations. However, the associations provided by this type of experiment are not conducive to finding causal relationships.

The goal of this manuscript is to highlight how large-scale experiments can impede the quest to uncovering the causal mechanism that determines the biology, and provide suggestions for addressing this problem through better experimental design and refining the biological question. We illustrate this on the MAPK pathway.
